Formalizing the Edmonds-Karp Algorithm

### 2016-03-07 paper/root.tex
Changes of meeting

\section{Introduction}
Computing the maximum flow of a network is an important problem in graph theory.
Many other problems, like maximum-bipartite-matching, edge-disjoint-paths, circulation-demand, as well as various scheduling and resource allocating problems can be reduced to it.
The Ford-Fulkerson method~\cite{FF56} describes a class of algorithms to solve the maximum flow problem, an important instance being the Edmonds-Karp algorithm~\cite{EK72},
which was one of the first algorithms to solve the maximum flow problem in polynomial time for the general case of real valued capacities.

In this paper, we present a formal verification of the Edmonds-Karp algorithm and its polynomial complexity bound.
The formalization is conducted entirely in the Isabelle/HOL proof assistant~\cite{NPW02}. 
Stepwise refinement techniques~\cite{Wirth71,Back78,BaWr98} allow us to elegantly structure our verification into an abstract proof of the Ford-Fulkerson method,
its instantiation to Edmonds-Karp algorithm, and finally its efficient implementation. The abstract parts of our verification closely follow the standard textbook presentation of Cormen et al.~\cite{CLRS09}. Being developed in the Isar~\cite{Wenzel99} proof language, our proofs are accessible even to non-Isabelle experts.

@@

\section{A short background on the Ford-Fulkerson method}\label{sec:background}
% \begin{comment}
%   GOAL: 
%     1) Remind the reader of FoFu. The educated reader should have the feeling:
%       Yes, now I recall FoFu, and understand (again) what it does.
%       
%     2) Set the field for further, more detailed descriptions in next section.
%   
% 
%   Short description of Network: Finite graph over nodes V, 
%     edge (u,v) annotated with capacity c(u,v).
%     Assuming distinct nodes s and t. 
%     Note on additional assumptions: 
%         No parallel edges, s only outgoing, t only incoming, all nodes on path from s to t. 
%         Can transform any Network to match these assumptions, preserving the flow cite{???}.
%         
%   s-t Flow: Annotation of edges with values, such that: 
%     1) Values smaller than capacities.
%     2) Kirchhoff law: sum of incoming flows = sum of outgoing flows, for all nodes except s t
%   
%   Value of flow: Incoming flow - outgoing flow of s
%   
%   Max-Flow problem. 
%     Rpt. why important?
%   
%   Cuts, intuition of min-cut >= max-flow
%   
%   Theorem: Min-Cut = Max-Flow
%   Proven via augmenting flow in residual graph. 
%     Present 1,2,3 of min-cut max-flow equivalences
%   
%   This yields an algorithm to compute max-flows, if we can compute flows in residual graph.
%     --> Simple way: Augmenting flow via augmenting path. ==> FoFu-scheme. 
%       Termination? In General: Only for integer capacities.
%       Edmonds-Karp: Shortest augmenting path. Always terminates! Complexity: O(|E||V|) outer loop iterations, BFS requires O|E|.
%   \end{comment}

In this section, we give a short introduction to the Ford-Fulkerson method, closely following the presentation in~\cite{CLRS09}.

@@

An \emph{flow} $f$ on a network is a labeling of the edges with real values satisfying the following constraints: 1) \emph{Capacity constraint}: the flow on each edge is a non-negative value smaller or equal to the edge's capacity; 2) \emph{Conservation constraint}: For all vertices except $s$ and $t$, the sum of flows over all incoming edges is equal to the sum of flows over all outgoing edges. 
The value of a flow $f$ is denoted by $|f|$, and defined to be the sum over the outgoing flows of $s$ minus the sum over the incoming flows of $s$.
Given a network $G$, the maximum flow problem is to find a flow with a maximum value among all flows of the network.

To simplify reasoning about the maximum flow problem, we assume that our network satisfies some additional constraints: 1) The source only has outgoing edges while the sink only has incoming edges; 2) If the flow network contains an edge $(u, v)$ then there is no \emph{parallel edge} $(v, u)$ in the reverse direction\footnote{With $u=v$, this also implies that there are no self loops.}; and 3) every vertex of the flow network must be on a path from $s$ to $t$. Note that any network can be transformed to a network with the aforementioned properties and the same maximum flow~\cite{CLRS09}.

@@

The Ford-Fulkerson method is a corollary of this theorem. It is based on a greedy approach: Starting from a zero flow, the value of the flow is iteratively increased until a maximum flow is reached. In order to increase the overall flow value, it may be necessary to redirect some flow, \ie to decrease the flow passed through specific edges. For this purpose the Ford-Fulkerson method defines the residual graph, which has edges both in parallel and reverse to the edges of the network. 
Each edge is labeled by the amount of flow that can be effectively moved along this edge, by either increasing or decreasing the flow on a network edge.
Formally, the residual graph $G_f$ of a flow $f$ is the graph induced by the 
edges with positive labels according to the following labeling function $c_f$:
\[ c_f (u, v) = 
  \begin{cases}
  c (u, v) - f(u, v) \hfill & \text{ if $(u, v) \in E$} \\
  f (v, u) \hfill & \text{ if $(v, u) \in E$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]

In each iteration, the Ford-Fulkerson method tries to find an \emph{augmenting path}, \ie a simple path from $s$ to $t$ in the residual graph, and then 
pushes as much flow as possible along this path to increase the value of the current flow. 
Formally, for an augmenting path $p$, one defines the \emph{residual capacity} $c_p$ as the minimum value over all edges of $p$:
\[c_f(p) = \min \{c_f(u, v): \text{$(u, v)$ is on  $p$}\}\]
An augmenting path yields a residual flow $f_p$, which is the flow that can be passed along this path:
\[ f_p (u, v) = 
  \begin{cases}
  c_f(p) \hfill & \text{ if $(u, v)$ is on $p$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]
\newcommand{\augment}{{\mathbin\uparrow}}
In order to actually push the flow induced by an augmenting path, we define the \emph{augment} function $f\augment f'$, which augments a flow $f$ in the network 
by any \emph{augmenting flow} $f'$, \ie any flow in the residual graph:
\[ (f \augment f') (u, v) = 
  \begin{cases}
  f (u, v) + f'(u, v) - f'(v, u) \hfill & \text{ if $(u, v) \in E$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]
Note that, for any edge in the network, the augmenting flow parallel to that edge is added to the flow, while the augmenting flow reverse to this edge is subtracted. 
This matches the intuition of passing flow in the indicated direction, by either increasing or decreasing the flow of an edge in the network.

@@

The Ford-Fulkerson method does not specify how to find an augmenting path in the residual graph, and there are several possible implementations with different execution times. The general method is only guaranteed to terminate for networks with rational capacities, while it may infinitely converge against non-maximal flows in the case of irrational edge capacities~\cite{FF56,Zwick95}. When always choosing a \emph{shortest} augmenting path, the number of iterations is bound by $O(VE)$, even for the general case of real valued capacities. Note that we write $V$ and $E$ instead of $|V|$ and $|E|$ for the number of nodes and edges if the intended meaning is clear from the context.
A shortest path can be found by breadth first search (BFS) in time $O(E)$, yielding the Edmonds-Karp algorithm~\cite{EK72} with an overall running time of $O(VE^2)$.

@@

In the following we present the corresponding formal proof in Isar:
\Snippet{augment_flow_presv_cap}
Its structure is exactly the same as the textbook proof, except that we had to also consider the case $(u,v)\notin E$, which is not mentioned in the informal proof at all, and easily discharged in our formal proof by the \isai{auto}-tactic after the \isai{qed}. We also use exactly the same justifications as the original proof, except that we had to use the fact that there are no parallel edges to show $c_f(v,u)=f(u,v)$, which is not mentioned in the original proof.

@@

The Isabelle Refinement Framework also supports data refinement, changing the representation of results according to a \emph{refinement relation}, 
which relates concrete with abstract results: Given a relation \isai{R}, \isai{\<Down>R m} is the largest set of concrete results that are related to an 
abstract result in \isai{m} by \isai{R}. If \isai{m=fail}, then also \isai{\<Down>R m = fail}.

@@

  Considerably more effort is required to show that the resulting algorithm terminates within $O(VE)$ iterations. 
  The idea of the proof is as follows: Edges that are reverse to edges on a shortest path cannot lie 
  on a shortest path itself.
  On every augmentation, at least one edge of the residual graph that lies on a shortest augmenting path is flipped. Thus, either the length of the shortest path increases,
  or the number of edges that lie on a shortest path decreases. As the length of a shortest path is at most $V$, there are no more than $O(VE)$ iterations.

@@

  \subsection{Network Checker}
  We also implemented an algorithm that takes as input a list of edges, a source node, and a target node.
  It converts these to a capacity matrix and an adjacency map, and checks whether the resulting graph satisfies our network assumptions.
  We proved that this algorithm returns the correct capacity matrix and adjacency map iff the input describes a valid network,
  and returns a failure value otherwise.

@@

  \begin{figure}[!h]
  \begin{tikzpicture}[thick,scale=.7, every node/.style={scale=1}] %change the scales if you like to reduce the size
    \begin{axis}[
      %title={Benchmark of the Edmonds-Karp algorithm},
        axis x line*= bottom,
        axis y line*= left,
        xmode = log,
        ymode = log,
        xlabel = Number of vertices in the network,
        ylabel = Execution time in milliseconds, %in milliseconds, modify the scale for lables
        xtick = {0,1000,...,5000},
%         ytick = data,
        xticklabels = {$1000$, $2000$, $3000$, $4000$, $5000$}, %corresponds to the first column in {SML, Sparse}
%         yticklabels = {$0.2$, $1.4$, $4.9$, $9.2$, $23.6$}, %corresponds to the second column in {SML, Sparse}
        smooth,
        cycle list name = black white,
        legend style = { 
          at={(0.59,0.4)}, 
            anchor=north west,
            draw=black, 
            fill= white,
            align=left
        }
    ]

    %start data for plotting
      \addplot table {
        1000 27   
        1500 124  
        2000 299  
        2500 665  
        3000 1335 
        3500 2526 
        4000 3111 
        4500 6917 
        5000 7743 
        5500 10452    
      }; \addlegendentry{SML, sparse};

      \addplot table {
        1000 16   
        1500 49   
        2000 168  
        2500 453  
        3000 842  
        3500 1524 
        4000 2136 
        4500 3438 
        5000 4734 
        5500 6369     
      }; \addlegendentry{Java, sparse};
    %end data for plotting
    \end{axis}
  \end{tikzpicture}
%
  \begin{tikzpicture}[thick,scale=.7, every node/.style={scale=1}] %change the scales if you like to reduce the size
    \begin{axis}[
      %title={Benchmark of the Edmonds-Karp algorithm},
        axis x line*= bottom,
        axis y line*= left,
        xmode = log,
        ymode = log,
        xlabel = Number of vertices in the network,
        ylabel = Execution time in milliseconds, %in milliseconds, modify the scale for lables
        xtick = {0,1000,1100,...,1400},
        ytick = {0,100,1000,10000},
        xticklabels = {$1000$, $1100$, $1200$, $1300$, $1400$}, %corresponds to the first column in {SML, Sparse}
%         yticklabels = {$0.2$, $1.4$, $4.9$, $9.2$, $23.6$}, %corresponds to the second column in {SML, Sparse}
        smooth,
        cycle list name = black white,
        legend style = { 
          at={(0.59,0.4)}, 
            anchor=north west,
            draw=black, 
            fill= white,
            align=left
        }
    ]

    %start data for plotting
      \addplot table {
        1000 3328 
        1050 3345 
        1100 4841 
        1150 5036 
        1200 5840 
        1250 6858 
        1300 7651 
        1350 8950 
        1400 8398 
        1450 10769     
      }; \addlegendentry{SML, dense};

      \addplot table {
        1000 4141 
        1050 4778 
        1100 5465 
        1150 6248 
        1200 7128 
        1250 8137 
        1300 9864 
        1350 11008
        1400 11902
        1450 13365     
      }; \addlegendentry{Java, dense};
    %end data for plotting
    \end{axis}
  \end{tikzpicture}

@@

  \caption{Edmonds-Karp benchmark}\label{fig:benchmark}
  \end{figure}

>>>>

\section{Introduction}
Computing the maximum flow of a network is an important problem in graph theory.
Many other problems, like maximum-bipartite-matching, edge-disjoint-paths, circulation-demand, as well as various scheduling and resource allocating problems can be reduced to it.
The Ford-Fulkerson method~\cite{FF56} describes a class of algorithms to solve the maximum flow problem. An important instance is the Edmonds-Karp algorithm~\cite{EK72},
which was one of the first algorithms to solve the maximum flow problem in polynomial time for the general case of networks with real valued capacities.

In this paper, we present a formal verification of the Edmonds-Karp algorithm and its polynomial complexity bound.
The formalization is conducted entirely in the Isabelle/HOL proof assistant~\cite{NPW02}. 
Stepwise refinement techniques~\cite{Wirth71,Back78,BaWr98} allow us to elegantly structure our verification into an abstract proof of the Ford-Fulkerson method,
its instantiation to the Edmonds-Karp algorithm, and finally an efficient implementation. The abstract parts of our verification closely follow the standard textbook presentation of Cormen et al.~\cite{CLRS09}. Being developed in the Isar~\cite{Wenzel99} proof language, our proofs are accessible even to non-Isabelle experts.

@@

\section{The Ford-Fulkerson Method}\label{sec:background}
% \begin{comment}
%   GOAL: 
%     1) Remind the reader of FoFu. The educated reader should have the feeling:
%       Yes, now I recall FoFu, and understand (again) what it does.
%       
%     2) Set the field for further, more detailed descriptions in next section.
%   
% 
%   Short description of Network: Finite graph over nodes V, 
%     edge (u,v) annotated with capacity c(u,v).
%     Assuming distinct nodes s and t. 
%     Note on additional assumptions: 
%         No parallel edges, s only outgoing, t only incoming, all nodes on path from s to t. 
%         Can transform any Network to match these assumptions, preserving the flow cite{???}.
%         
%   s-t Flow: Annotation of edges with values, such that: 
%     1) Values smaller than capacities.
%     2) Kirchhoff law: sum of incoming flows = sum of outgoing flows, for all nodes except s t
%   
%   Value of flow: Incoming flow - outgoing flow of s
%   
%   Max-Flow problem. 
%     Rpt. why important?
%   
%   Cuts, intuition of min-cut >= max-flow
%   
%   Theorem: Min-Cut = Max-Flow
%   Proven via augmenting flow in residual graph. 
%     Present 1,2,3 of min-cut max-flow equivalences
%   
%   This yields an algorithm to compute max-flows, if we can compute flows in residual graph.
%     --> Simple way: Augmenting flow via augmenting path. ==> FoFu-scheme. 
%       Termination? In General: Only for integer capacities.
%       Edmonds-Karp: Shortest augmenting path. Always terminates! Complexity: O(|E||V|) outer loop iterations, BFS requires O|E|.
%   \end{comment}

In this section, we give a short introduction to the Ford-Fulkerson method, closely following the presentation by Cormen et al.~\cite{CLRS09}.

@@

A \emph{flow} $f$ on a network is a labeling of the edges with real values satisfying the following constraints: 1) \emph{Capacity constraint}: the flow on each edge is a non-negative value smaller or equal to the edge's capacity; 2) \emph{Conservation constraint}: For all vertices except $s$ and $t$, the sum of flows over all incoming edges is equal to the sum of flows over all outgoing edges. 
The value of a flow $f$ is denoted by $|f|$, and defined to be the sum over the outgoing flows of $s$ minus the sum over the incoming flows of $s$.
Given a network $G$, the maximum flow problem is to find a flow with a maximum value among all flows of the network.

To simplify reasoning about the maximum flow problem, we assume that our network satisfies some additional constraints: 1) The source only has outgoing edges while the sink only has incoming edges; 2) If the network contains an edge $(u, v)$ then there is no \emph{parallel edge} $(v, u)$ in the reverse direction\footnote{With $u=v$, this also implies that there are no self loops.}; and 3) every vertex of the network must be on a path from $s$ to $t$. Note that any network can be transformed to a network with the aforementioned properties and the same maximum flow~\cite{CLRS09}.

@@

The Ford-Fulkerson method is a corollary of this theorem. It is based on a greedy approach: Starting from a zero flow, the value of the flow is iteratively increased until a maximum flow is reached. In order to increase the overall flow value, it may be necessary to redirect some flow, \ie to decrease the flow passed through specific edges. For this purpose the Ford-Fulkerson method defines the residual graph, which has edges both in parallel and reverse to the edges of the network. 
Each edge is labeled by the amount of flow that can be effectively passed along this edge, by either increasing or decreasing the flow on a network edge.
Formally, the residual graph $G_f$ of a flow $f$ is the graph induced by the 
edges with positive labels according to the following labeling function $c_f$:
\[ c_f (u, v) = 
  \begin{cases}
  c (u, v) - f(u, v) \hfill & \text{ if $(u, v) \in E$} \\
  f (v, u) \hfill & \text{ if $(v, u) \in E$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]

In each iteration, the Ford-Fulkerson method tries to find an \emph{augmenting path}, \ie a simple path from $s$ to $t$ in the residual graph, and then 
pushes as much flow as possible along this path to increase the value of the current flow. 
Formally, for an augmenting path $p$, one first defines the \emph{residual capacity} $c_p$ as the minimum value over all edges of $p$:
\[c_f(p) = \min \{c_f(u, v): \text{$(u, v)$ is on  $p$}\}\]
An augmenting path then yields a residual flow $f_p$, which is the flow that can be passed along this path:
\[ f_p (u, v) = 
  \begin{cases}
  c_f(p) \hfill & \text{ if $(u, v)$ is on $p$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]
\newcommand{\augment}{{\mathbin\uparrow}}%
Finally, to actually push the flow induced by an augmenting path, we define the \emph{augment} function $f\augment f'$, which augments a flow $f$ in the network 
by any \emph{augmenting flow} $f'$, \ie any flow in the residual graph:
\[ (f \augment f') (u, v) = 
  \begin{cases}
  f (u, v) + f'(u, v) - f'(v, u) \hfill & \text{ if $(u, v) \in E$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]
Note that, for any edge in the network, the augmenting flow parallel to that edge is added to the flow, while the augmenting flow reverse to this edge is subtracted. 
This matches the intuition of passing flow in the indicated direction, by either increasing or decreasing the flow of an edge in the network.

@@

The Ford-Fulkerson method does not specify how to find an augmenting path in the residual graph. There are several possible implementations with different execution times. The general method is only guaranteed to terminate for networks with rational capacities, while it may infinitely converge against non-maximal flows in the case of irrational edge capacities~\cite{FF56,Zwick95}. When always choosing a \emph{shortest} augmenting path, the number of iterations is bound by $O(VE)$, even for the general case of real valued capacities. Note that we write $V$ and $E$ instead of $|V|$ and $|E|$ for the number of nodes and edges if the intended meaning is clear from the context.
A shortest path can be found by breadth first search (BFS) in time $O(E)$, yielding the Edmonds-Karp algorithm~\cite{EK72} with an overall running time of $O(VE^2)$.

@@

In the following we present the corresponding formal proof in Isar:
\Snippet{augment_flow_presv_cap}
The structure of the Isar proof is exactly the same as that of the textbook proof, except that we had to also consider the case $(u,v)\notin E$, which is not mentioned in the informal proof at all, and easily discharged in our formal proof by the \isai{auto}-tactic after the \isai{qed}. We also use exactly the same justifications as the original proof, except that we had to use the fact that there are no parallel edges to show $c_f(v,u)=f(u,v)$, which is not mentioned in the original proof.

@@

The Isabelle Refinement Framework also supports data refinement, changing the representation of results according to a \emph{refinement relation}, 
which relates concrete with abstract results: Given a relation \isai{R}, \isai{\<Down>R m} is the largest set of concrete results that are related to an 
abstract result in \isai{m} by \isai{R}. If \isai{m = fail}, then also \isai{\<Down>R m = fail}.

@@

  Considerably more effort is required to show that the resulting algorithm terminates within $O(VE)$ iterations. 
  The idea of the proof is as follows: Edges that are reverse to edges on a shortest path cannot lie 
  on a shortest path itself.
  On every augmentation, at least one edge of the residual graph that lies on a shortest augmenting path is flipped. Thus, either the length of the shortest path increases,
  or the number of edges that lie on some shortest path decreases. As the length of a shortest path is at most $V$, there are no more than $O(VE)$ iterations.
  % TODO: Add more explanation here, like: DIstinguish two cases: 1) ... 2) The distance s-t in cf is still the same, but then ...

@@

  \subsection{Network Checker}
  Additionally, we implemented an algorithm that takes as input a list of edges, a source node, and a target node.
  It converts these to a capacity matrix and an adjacency map, and checks whether the resulting graph satisfies our network assumptions.
  We proved that this algorithm returns the correct capacity matrix and adjacency map iff the input describes a valid network,
  and returns a failure value otherwise.

@@

  \begin{figure}[!h]
  \begin{tikzpicture}[thick,scale=.8, every node/.style={scale=1}] %change the scales if you like to reduce the size
    \begin{axis}[
      %title={Benchmark of the Edmonds-Karp algorithm},
        axis x line*= bottom,
        axis y line*= left,
        xmode = log,
        ymode = log,
%         xlabel = Number of vertices in the network,
%         ylabel = Execution time in seconds, 
        xtick = {0,1000,...,5000},
        ytick = {0.01,0.1,1,10},
        xticklabels = {$1000$, $2000$, $3000$, $4000$, $5000$},
        yticklabels = {$0.01$,$0.1$,$1$,$10$}, 
        smooth,
        cycle list name = black white,
        legend style = { 
          at={(0.59,0.4)}, 
            anchor=north west,
            draw=black, 
            fill= white,
            align=left
        }
    ]

    %start data for plotting
      \addplot table {
        1000 .027   
        1500 .124  
        2000 .299  
        2500 .665  
        3000 1.335 
        3500 2.526 
        4000 3.111 
        4500 6.917 
        5000 7.743 
        5500 10.452    
      }; \addlegendentry{SML, sparse};

      \addplot table {
        1000 .016   
        1500 .049   
        2000 .168  
        2500 .453  
        3000 .842  
        3500 1.524 
        4000 2.136 
        4500 3.438 
        5000 4.734 
        5500 6.369     
      }; \addlegendentry{Java, sparse};
    %end data for plotting
    \end{axis}
  \end{tikzpicture}
%
  \begin{tikzpicture}[thick,scale=.8, every node/.style={scale=1}] %change the scales if you like to reduce the size
    \begin{axis}[
      %title={Benchmark of the Edmonds-Karp algorithm},
        axis x line*= bottom,
        axis y line*= left,
        xmode = log,
        ymode = log,
%         xlabel = Number of vertices in the network,
%         ylabel = Execution time in seconds, 
        xtick = {0,1000,1100,...,1400},
        ytick = {2,4,...,10},
        xticklabels = {$1000$, $1100$, $1200$, $1300$, $1400$}, 
        yticklabels = {$2$, $4$,$6$,$8$,$10$}, 
        smooth,
        cycle list name = black white,
        legend style = { 
          at={(0.59,0.4)}, 
            anchor=north west,
            draw=black, 
            fill= white,
            align=left
        }
    ]

    %start data for plotting
      \addplot table {
        1000 3.328 
        1050 3.345 
        1100 4.841 
        1150 5.036 
        1200 5.840 
        1250 6.858 
        1300 7.651 
        1350 8.950 
        1400 8.398 
        1450 10.769     
      }; \addlegendentry{SML, dense};

      \addplot table {
        1000 4.141 
        1050 4.778 
        1100 5.465 
        1150 6.248 
        1200 7.128 
        1250 8.137 
        1300 9.864 
        1350 11.008
        1400 11.902
        1450 13.365     
      }; \addlegendentry{Java, dense};
    %end data for plotting
    \end{axis}
  \end{tikzpicture}

@@

%   \center{\footnotesize The x-axis shows the number of nodes, the y axis the execution time in seconds.}
  \caption{Benchmark of different implementations. The x-axis shows the number of nodes, the y axis the execution time in seconds.}\label{fig:benchmark}
  \end{figure}
